---
title: "Automation_Report"
format: html

---
title: "Predicting Future Employment in the U.S. Due to Automation"
format: html


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
```

# Introduction

Automation is rapidly transforming the global labor market. From robotic/industrial automation in manufacturing to AI usage in administrative roles. This report evaluates how automation affects future U.S. employment growth using Bureau of Labor Statistics (BLS) projections and a basic predictive model.

# Data Sources

- **BLS Employment Projections (2023–2033):**  
  https://www.bls.gov/emp/tables/emp-by-detailed-occupation.htm

- **BLS Industry Employment (2013–2033):**  
  https://www.bls.gov/emp/tables/employment-by-major-industry-sector.htm

- **U.S. Census Bureau: ASM Robotics Supplement (2018–2021):**  
  https://www.census.gov/data/experimental-data-products/robotics.html

- **Brookings Automation Risk Estimates:**  
  https://www.brookings.edu/interactives/automation-and-artificial-intelligence-the-future-of-work/

# Industry Employment Trends (2013–2033)

```{r}
industry_data <- read_excel("industry.xlsx", sheet = "Table 2.1", skip = 1)
colnames(industry_data)[1] <- "Industry"

selected_industries <- c("Manufacturing", "Retail trade", "Transportation and warehousing", "Administrative and support services")

industry_data_filtered <- industry_data %>%
  filter(Industry %in% selected_industries) %>%
  select(Industry, `Employment, 2013`, `Employment, 2023`, `Employment, 2033`) %>%
  pivot_longer(cols = starts_with("Employment"),
               names_to = "Year",
               values_to = "Employment") %>%
  mutate(Year = parse_number(Year),
         Employment = as.numeric(Employment))

ggplot(industry_data_filtered, aes(x = Year, y = Employment, color = Industry)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Industry Employment Trends (2013–2033)",
       subtitle = "Source: BLS Table 2.1",
       x = "Year", y = "Employment (thousands)") +
  theme_minimal()
```

This graph shows declining or stagnant growth in manufacturing and retail — sectors known for high automation risk — while transportation and warehousing show modest growth.

```{r}
occupation_data <- read_excel("occupation (1).xlsx", sheet = "Table 1.1", skip = 1)
colnames(occupation_data)[1:4] <- c("Occupation", "SOC", "Employment_2023", "Employment_2033")

occupation_cleaned <- occupation_data %>%
  filter(!is.na(SOC) & !is.na(Employment_2023)) %>%
  mutate(
    Employment_2023 = as.numeric(Employment_2023),
    Employment_2033 = as.numeric(Employment_2033),
    Growth = Employment_2033 - Employment_2023,
    GrowthRate = Growth / Employment_2023
  ) %>%
  filter(Employment_2023 > 0)
```

We calculate a new variable `GrowthRate`, which represents the expected change in employment from 2023 to 2033. This will be used as the target in our predictive model.

```{r}
robotics <- read_excel("ASM Robotics.xlsx", skip = 3) %>%
  rename_with(~str_trim(.x)) %>%
  rename(Industry = 2) %>%
  slice(-1) %>%
  select(Industry, 
         `2018 Percent of Plants with Robots`,
         `2019 Percent of Plants with Robots`,
         `2020 Percent of Plants with Robots`,
         `2021 Percent of Plants with Robots`) %>%
  filter(Industry %in% c("Manufacturing", 
                         "Food manufacturing", 
                         "Transportation equipment manufacturing", 
                         "Computer and electronic product manufacturing")) %>%
  pivot_longer(cols = -Industry, names_to = "Year", values_to = "Percent") %>%
  mutate(
    Year = parse_number(Year),
    Percent = as.numeric(Percent)
  )

# Now plot
ggplot(robotics, aes(x = Year, y = Percent * 100, color = Industry)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(0, 100)) +
  scale_x_continuous(breaks = 2018:2021) +
  labs(
    title = "Robot Adoption in U.S. Manufacturing Industries (2018–2021)",
    subtitle = "Percent of Plants Using Robots",
    x = "Year",
    y = "Percent of Plants (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

```

> **Annotation:** Industries like *Computer & Electronics* and *Transportation Equipment* lead robot integration, reflecting deeper automation penetration.

------------------------------------------------------------------------

# Employment Projection With and Without Automation

We assign each occupation group an estimated `AutomationRisk` score based on information from Brookings. This feature will allow us to simulate how automation exposure contributes to job growth or decline.

```{r}
occupation_data <- read_excel("occupation (1).xlsx", sheet = "Table 1.1", skip = 1)
colnames(occupation_data)[1:4] <- c("Occupation", "SOC", "Employment_2023", "Employment_2033")
occupation_data$Occupation <- str_trim(occupation_data$Occupation)

automation_risk <- tribble(
  ~Occupation, ~AutomationRisk,
  "Management occupations", 0.2,
  "Business and financial operations occupations", 0.3,
  "Computer and mathematical occupations", 0.2,
  "Architecture and engineering occupations", 0.35,
  "Life, physical, and social science occupations", 0.25,
  "Community and social service occupations", 0.4,
  "Legal occupations", 0.3,
  "Educational instruction and library occupations", 0.35,
  "Arts, design, entertainment, sports, and media occupations", 0.4,
  "Healthcare practitioners and technical occupations", 0.15,
  "Sales and related occupations", 0.7,
  "Office and administrative support occupations", 0.8,
  "Food preparation and serving related occupations", 0.85,
  "Transportation and material moving occupations", 0.75
)

merged <- occupation_data %>%
  filter(Occupation %in% automation_risk$Occupation) %>%
  mutate(across(starts_with("Employment"), as.numeric)) %>%
  left_join(automation_risk, by = "Occupation") %>%
  mutate(Adjusted_2033 = Employment_2033 * (1 - AutomationRisk))

plot_df <- merged %>%
  select(Occupation, Employment_2023, Employment_2033, Adjusted_2033) %>%
  pivot_longer(cols = starts_with("Employment") | starts_with("Adjusted"),
               names_to = "Scenario", values_to = "Employment") %>%
  mutate(Scenario = recode(Scenario,
                           "Employment_2023" = "2023",
                           "Employment_2033" = "2033 (Projected)",
                           "Adjusted_2033" = "2033 (Adj. for Automation)"))

ggplot(plot_df, aes(x = Occupation, y = Employment, fill = Scenario)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Employment Projection With vs Without Automation Adjustment", y = "Employment (in thousands)", x = NULL) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 60,hjust = 1))
```

------------------------------------------------------------------------

# Automation Risk

```{r}

occupation_cleaned <- occupation_cleaned %>%
  left_join(automation_risk, by = "Occupation") %>%
  filter(!is.na(AutomationRisk))
```

------------------------------------------------------------------------

# Predictive Model Using Linear Regression

```{r}
lm_model <- lm(GrowthRate ~ Employment_2023 + AutomationRisk, data = occupation_cleaned)
occupation_cleaned$Predicted_LM <- predict(lm_model, newdata = occupation_cleaned)
```

This linear model predicts `GrowthRate` based on current employment and automation exposure. As expected, jobs with high automation risk tend to show negative growth.

------------------------------------------------------------------------

# Visualization: Actual vs Predicted Growth Rate

```{r}
ggplot(occupation_cleaned, aes(x = GrowthRate, y = Predicted_LM, label = Occupation)) +
  geom_point(color = "blue", size = 3) +
  geom_text(size = 3, hjust = 0.5, vjust = -1.2, check_overlap = TRUE) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Predicted vs Actual Growth Rate (2023–2033)",
       subtitle = "Linear Model Using Employment + Automation Risk",
       x = "Actual Growth Rate", y = "Predicted Growth Rate") +
  theme_minimal()
```

This plot shows strong correlation between automation risk and future job contraction. Occupations above the red line are under-predicted, while those below may be overestimated by the model.

------------------------------------------------------------------------

# Key Takeaways

-   **Automation risk is a reliable predictor of job decline.** High-risk occupations are growing slower or shrinking.
-   **Human-centered roles** like healthcare, education, and science remain more resilient

------------------------------------------------------------------------

