---
title: "dogR (R package)"
format: html
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
#devtools::load_all()
```


The goal of `dogR` is to explore a dog dataset using functions to learn some fun characteristics and predictive modeling.

## Installation

You can install the development version of dogR from [GitHub](https://github.com/) with:

``` r
install.packages("devtools")
devtools::install_github("ADC-405-S26/dogR")
```
## Warning
When loading into another project, ensure that all other packages are cleared or updated when running.

## Example

Below are basic examples which shows you how to use the package's functions:

Load the library
```{r example}
library(dogR)
```

Load the data set
```{r}
dogs_dataset <- dogs_dataset
head(dogs_dataset)
```

Using get_breed_summary(x), you can get a quick summary of a specific dog breed.
```{r summary}
get_breed_summary("Airedale Terrier")
```
filter_dogs() takes in multiple filtering variables and returns an updated table.
```{r filter}
small_young_dogs <- filter_dogs(
  max_weight = 10, 
  max_age = 3, 
  gender = "Female"
)
head(small_young_dogs)
```

plot_breed_weight("") returns a weight distribution count for a specific breed
```{r plot}
plot_breed_weight("Poodle")
```
